---
title: "Divvy Capstone"
author: "Lanai Toledo"
date: "2023-07-06"
output:
  pdf_document: default
  html_document:
    df_print: paged

---
```{r echo= FALSE}

library(tinytex)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Google Data Analytics Capstone Project: How does a bike-share navigate speedy success?
 
## About the Company
#### In 2016, Cyclistic launched a successful bike-share offering having a fleet of 5,824 bicycles that are tracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system at any time.
#### Members are riders who have an annual subscription while casual riders are users who are single-ride or full-day pass and do not have an annual membership plan.
#### The director of marketing is looking to maximize the number of annual memberships as they are more profitable than single-ride or full-day passes and wants to design marketing strategies to entice casual riders to become annual members. 

## The questions that need to be answered are:
#### How do annual members and casual riders use Cyclistic bikes differently?
#### Why would casual riders buy Cyclistic annual memberships?
#### How can Cyclistic use digital media to influence casual riders to become members?

## Gather the Data
#### I will be using the public dataset located <https://divvy-tripdata.s3.amazonaws.com/index.html> . The data has been made available by Motivate International Inc. under this [license]: <https://www.divvybikes.com/data-license-agreement> . 
#### Key Tasks Followed:
##### Downloaded data and locally stored copies.
##### Data is downloaded for the year 2022 and includes January through December.
##### The data is in CSV (comma-separated values) format, and there are a total of 13 columns.


### Set Up Environment
###### Load packages in R to enable cleaning and data transformation

```{r environment}

library(tidyverse)  
library(lubridate)  
library(ggplot2)
getwd() 
```

### Upload data files as CSV

```{r load data}
jan <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202201-divvy-tripdata.csv")
feb <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202202-divvy-tripdata.csv")
mar <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202203-divvy-tripdata.csv")
apr <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202204-divvy-tripdata.csv")
may <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202205-divvy-tripdata.csv")
jun <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202206-divvy-tripdata.csv")
jul <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202207-divvy-tripdata.csv")
aug <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202208-divvy-tripdata.csv")
sep <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202209-divvy-publictripdata.csv")
oct <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202210-divvy-tripdata.csv")
nov <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202211-divvy-tripdata.csv")
dec <- read.csv("C:/Users/toled/OneDrive/Desktop/Case Study 1/2022 CSV/202212-divvy-tripdata.csv")
```

### Inspect data columns before combining into single dataframe

```{r column inspect}
colnames(jan)
colnames(feb)
colnames(mar)
colnames(apr)
colnames(may)
colnames(jun)
colnames(jul)
colnames(aug)
colnames(sep)
colnames(oct)
colnames(nov)
colnames(dec)
```

### Inspect dataframes to make sure strucure is the same before combining

```{r inspect structure}
str(jan)
str(feb)
str(mar)
str(apr)
str(may)
str(jun)
str(jul)
str(aug)
str(sep)
str(oct)
str(nov)
str(dec)
```

### Combine all months data into one dataframe

```{r combine dataframes}
all_trips <- bind_rows(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)
```

### Inspect the new table that has been created

```{r inspect new table}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)
```

### Add columns that list the date, month, day, and year of each ride
###### This will allow us to aggregate ride data for each month, day, or year 

```{r date}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

### Add a "ride_length" calculation to all_trips (in seconds)

```{r ride_length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

### Review the structure of all the columns that were just created

```{r}
str(all_trips)
```

### Convert "ride_length"  to numeric so we can run calculations on the data

```{r convert ride_length}
is.numeric(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### Ensure all "bad" data is removed (bikes taken out of docks for inspection and neg ride_length)
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

### Descriptive analysis on ride_length (all figures in seconds)
```{r ride_length analyze}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride

summary(all_trips_v2$ride_length)
```

### Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

### See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

### Reorder the days of the week
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### See average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

### Analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n()							
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday)
```

### Visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

```{r visualize by month}
all_trips_v2 %>% 
  group_by(member_casual, month)%>%
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Create a csv file that we can be further analyzed in Tableau

```{r include=FALSE}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "C:/Users/toled/OneDrive/Desktop/Case Study 1/avg_ride_length.csv")

write.csv(all_trips, file= "C:/Users/toled/OneDrive/Desktop/Case Study 1/all_trips.csv")
```

## Summary of Analysis 
#### Casual riders tend to take fewer trips than members.
#### Casual riders tend to take longer trips than members.
#### Casual riders take more rides on weekends than during the week.
#### Casual ridership is higest during the summer months. 

## Recommendations
#### Offer discounts on annual membership during spring and summer months to entice causl riders into becoming annual members.
#### Increasing weekend single ride and singe day pricing to incentivize membership.
#### Increasing cost for trips lasting more than a certain time for non-members to incentivize membership.
